ESRC SDAI Application - Understanding domestic abuse using Big Data

### 2019-07-11 JeS_fields.tex
Anna crappy summary

Who are the people most vulnerable to becoming victims of domestic abuse? To what extent do neighbourhood characteristics predict the prevalence of domestic abuse? How can we prevent serious harm and what affects victims' decision to finally report the abuse? What is the quantifiable, long-term impact of domestic abuse victimisation, or witnessing domestic abuse as a child on socio-economic outcomes in later life? How do changes in alcohol consumption and financial stability affect the prevalence of domestic abuse?

Addressing these questions are fundamental for a deeper understanding of the constellation of individual and environmental factors perpetuating domestic abuse, and the long-term consequences it can have on the lives of those affected. Designing effective policy interventions to protect victims can only be based on a sound understanding of the type of environments where domestic abuse is likely to occur. In this research programme, we combine various data sources to provide a comprehensive answer to these questions in the context of the UK.

Domestic abuse is increasingly recognised as a serious public health concern in many countries, including the UK. It is estimated that in the year ending 2018 March, 2 million adults in England and Wales (7.9\% of women and 4.2\% of men) have experienced some form of domestic abuse \cite{ONS}. Escalation of domestic abuse can result in homicide, and the risk of lethal violence is especially high in the case of female victims: 41\% of femicides, and 0.6\% of homicides committed in the year ending 2018 March were domestic abuse-related in England and Wales \cite{homic}. The long-lasting effect of domestic abuse are not limited to the direct target of abuse, as it can have severe developmental impacts on children witnessing the abuse. Taking into account the emotional and physical consequences of the abuse suffered, the economic cost of domestic abuse in England and Wales between April, 2016 and March, 2017 was estimated to be \pounds 66 billion \cite{costs}.

Despite the vast economic costs associated with it, there is relatively little academic research focusing on domestic abuse in the context of the UK. Studies using police-recorded domestic abuse data have limited ability to derive useful insights about this phenomenon due to the vast levels of underreporting. In contrast, studies using qualitative methods have limited ability to derive useful insights about this phenomenon.

@@

In the UK, the most reliable source of data on domestic abuse is the Crime Survey for England and Wales (CSEW), an annual, cross-sectional representative victimisation survey across England and Wales, collecting data from 30,000--40,000 households every year \cite{OfficeforNationalStatistics2019}. The survey has a self-completion module containing questions about the respondent's experiences of domestic abuse throughout their life (since the age of 16), and in the past 12 months in particular.  While the Office for National Statistics (ONS) regularly publishes findings from the CSEW, these are mostly exploratory and descriptive in nature. In spite of the rich information provided by this survey, only a limited number of studies have utilised the CSEW (e.g., \citeNP{Khalifeh2013a}) to understand domestic abuse.

 By recognising its potential, we will rely on the CSEW and combine in with other external datasets to examine various aspects on domestic abuse in England and Wales, and provide a better understanding of this phenomenon. We will explore the individual- and environmental-level risk factors of domestic abuse, including the demographic and socio-economic predictors of victimisation, as well as the neighbourhood characteristics affecting it. Drawing on previous reports on police mis-recording of domestic abuse across England, we will explore whether this affects specific victim groups disproportionately, and its consequences regarding victim's trust in the Criminal Justice System. By combining the rich information in the CSEW with a police crime dataset provided by the West Midlands Police, we will explore the victim- and perpetrator-specific predictors of domestic abuse resulting serious harm and the time-varying and time-invariant predictors of reporting the abuse. The CSEW will also allow us to estimate the causal effect of direct and indirect domestic abuse victimisation on socio-economic and health-related outcomes in adulthood and the likelihood of engaging in violent criminal behaviours in childhood. In addition, we will explore how time-varying external factors, such as changes in alcohol consumption and financial health affect the reported number of domestic abuse incidents. Our previous investigation have revealed that the reported number of alcohol-related domestic abuse increases by 61\% following the victory of the England national football team, highlighting the profound effect external events can have on perpetrators' propensity to engage in violence.

para on implications

>>>>

Who are the people most vulnerable to becoming victims of domestic abuse? To what extent do neighbourhood characteristics predict the prevalence of domestic abuse? What predicts serious harm and victims' decision to finally report the abuse? What is the quantifiable, long-term impact of domestic abuse victimisation, or witnessing domestic abuse as a child on socio-economic outcomes in later life? How do changes in alcohol consumption and financial stability affect the prevalence of domestic abuse?

Addressing these questions is fundamental for a deeper understanding of the far-reaching consequences of domestic abuse victimisation and the individual and environmental factors precipitating it. This knowledge can in turn form the basis of effective policy interventions aiming to encourage willingness to report and safeguarding groups vulnerable to domestic abuse. In this research programme, we aim to provide a comprehensive answer to these (and other) questions related to domestic abuse, with a series of quantitative analyses drawing on various data sources.

It is estimated that in the year ending 2018 March, 2 million adults in England and Wales (7.9\% of women and 4.2\% of men) have experienced some form of domestic abuse \cite{ONS}. Domestic abuse can result in homicide, and the risk of lethal violence is especially high in the case of female victims: 41\% of femicides, and 0.6\% of homicides committed in the year ending 2018 March were domestic abuse-related \cite{homic}. The long-lasting effect of domestic abuse are not limited to the direct target of abuse, as it can have severe developmental impacts on children witnessing the abuse. The economic cost of domestic abuse in England and Wales between April, 2016 and March, 2017 was estimated to be \pounds 66 billion \cite{costs}.

Despite the extent of the problem, there is relatively little academic research focusing on domestic abuse in the context of the UK, perhaps due to the lack of data sources. Using police data is problematic due to the vast levels of underreporting, while insights from qualitative approaches are hard to generalise. In the UK, the most reliable source of data on domestic abuse is the Crime Survey for England and Wales (CSEW), an annual, cross-sectional representative survey collecting information on experiences of domestic abuse \cite{OfficeforNationalStatistics2019}.  While the Office for National Statistics (ONS) regularly publishes findings from the CSEW, these are mostly exploratory and descriptive in nature. In spite of the rich information provided by this survey, only a limited number of studies have utilised the CSEW (e.g., \citeNP{Khalifeh2013a}) to understand domestic abuse.

@@

Our project takes advantage of the rich information contained in the CSEW, and draws on other data sources to characterise domestic abuse in England and Wales. We will identify the individual and neighbourhood-level predictors of domestic abuse victimisation, and the factors affecting victims' willingness to contact the police. We will also explore whether there is any evidence that the police's response depend on the individual victim characteristics. We will quantify the causal effect of domestic abuse on the behavioural outcomes of child and the socio-economic outcomes of adult victims. Finally, we will explore how alcohol consumption, spending on gambling, and changes in benefit receipts affect the prevalence of domestic abuse.

para on implications (500 chars)
